### 2019-12-31 manuscript/main.tex
finished results section

While historically the phylogenetic informativeness of the \emph{pol} region of the HIV genome was initially contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic analysis and cluster inference, often due to the availability of \emph{pol} sequences from routine clinical genotyping by commercial Sanger sequencing. In a recent meta study, Novitsky \emph{et al.} (forthcoming) found that 102 out of 107 published studies of HIV cluster inference between 2016 and 2019 used the \emph{pol} region alone.

@@

\subsection*{Profile sampling captures within-host diversity}

We examine how much

Profile sampling allows us to visualize the degree of polymorphism within a patient (Figure 1) and sample genomic sequences in proportion to the within-host variation. It can also build a consensus sequence for easy comparison of new approaches to established methods. By visualizing the degree of polymorphism, we can assess whether genomic variation exists at significantly high enough frequencies such that neither consensus genomes nor any other point estimator are sufficient summaries.

Figure 1 shows substantial variation exists in the HIV profiles of all patients. Across	all patients, we found that 5.63 percent of sites in the HIV genome are polymorphic, with 11.4 percent of those sites in the gag region, 21.4 percent of those sites in the pol region, and 29.5 percent of those sites in the env region of the genome. Here we define polymorphism to mean that a particular site has greater than 1% variation. This falls within the range of prior studies on within-host variation \parencite{li, zanini}. This suggests that consensus genomes, and in particular Sanger sequencing of the pol region, discards variation that is informative for inferring the phylogeny.

\subsection*{Phylogenetic estimates are sensitive to inter-host diversity}

Given that previous studies on individuals infected with HIV have shown high intra-patient variation, and given that we found similar results in our set of data (Figure 1), we investigated its impact on downstream topological variation in the phylogenetic tree.  The results show that trees inferred from the profile-sampled sequences do not cluster around any of the trees inferred from a genome point estimate (Figure 2).

When and MDS plot is generated for only the profile-sampled trees (Figure 3), they appeared to form two clusters. This is likely due to zoom on the profile-sampled trees, as can be seen from the axes scales for Figure 2 and Figure 3. Supertrees of the two clusters reveal that most of the clades in the two subtrees are the same, with only patients out of 37 having a different topological placement on the two trees (Figure 4). This suggests that distinct clusters of phylogenetic trees exist given the intra-patient variation, but that the differences between them are not large.  However, the consensus genome appears to be an inadequate estimate summary for the intra-patient variation, as can be seen in Figure 5.

@@

@@

@@

Our construction of HIV profiles from NGS data is limited by the accuracy of the NGS assays themselves. The codon frequencies in the profiles may be biased measures of the true within-host diversity because of biases in PCR amplification during sample preparation. Sequencing protocols such as Primer ID \cite{jabara} have been introduced to reduce and correct for these biases. Unfortunately, the Primer ID protocol supports only a limited region of the HIV genome, and has not yet been extended to support near-whole genome sequencing of HIV, but future work could study this limited region to determine if cluster inference changes with the implementation of a Primer ID protocol.

The true HIV transmission network is unknown, but phylogenetic analysis and cluster inference are promising tools for aiding clinicians and public health officials \cite{fauci}. Current phylogenetic approaches do not fully utilize the information on within-host diversity available from deeper and longer NGS sequencing technologies. As NGS data sets are increasingly available and more fully sample from the current HIV epidemic,

>>>>

While historically the phylogenetic informativeness of the \emph{pol} region of the HIV genome was initially contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic analysis and cluster inference, often due to the availability of \emph{pol} sequences from routine clinical genotyping by commercial Sanger sequencing. In a meta study of published studies on HIV cluster infereance, 98 out of 105 (93\%) analyzed the \emph{pol} region \cite{hassan}.

@@

\subsection*{Profile sampling estimates within-host diversity}

We estimate the within-host diversity for individuals as the average percent difference across all pairwise comparisons of their 500 profile-sampled nucleotide sequences. These pairwise differences are calculated using the Hamming distance \parencite{allam} (which is also called the $p$-distance in the literature on HIV genetic diversity \parencite{maldarelli, hassan}). Figure 1 shows the estimated percent diversity in each region across individuals, ordered by \emph{env}, which we expected \emph{a priori} to be the most variable region. The largest estimated diversity is in \emph{env} for individual MC28 (4.0\%), and \emph{env} has the largest range in estimated diversity (0.2\% to 4.0\%). The other regions have ranges of 0.2\% to 1.9\% (prrt), 0.1\% to 2.0\% (\emph{int}) and 0.2\% to 2.6\% (wgs). These ranges are comparable to prior studies on within-host diversity \parencite{li, zanini}. These estimates of within-host diversity from the HIV profiles suggest that the consensus approach is discarding potentially-relevant information that is present in the NGS data.

\subsection*{Phylogenetic estimates are sensitive to within-host diversity}

Next, we investigate the impact of within-host diversity on phylogenetic topology and evolutionary distance estimates. To examine the variation in topology, we calculate the pairwise geodesic distance \parencite{billera,owen} among the 500 phylogenies from the profile samples, as well as the phylogenies from the NGS and Sanger consensus sequences. Then we perform multi-dimensional scaling (MDS) on the resulting distance matrix to visualize the topological space in two dimensions (Figure 2). The results show that the topolical space is multi-model, and the consensus phylogenies inferred from the consensus sequences do not fully summarize the clustering that occurs in topological space. The prrt region displays four topological clusters, and the NGS and Sanger consensus phylogenies are located in difference clusters. There are two clusters in \emph{int}, and the NGS consensus phylogeny lies at the center of one of the clusters, while the Sanger consensus phylogeny is an outlier. The consensus phlyogenies lie at opposite ends of the primary MDS axis in \emph{env}, and of the secondary MDS axis in the wgs region.

To examine the variation in estimated evolutionary distance, we sum the branch lengths within each phylogeny across all branches and across only the tip branches. We visualize the distribution of these branch length sums in Figure 3. Overall, the estimates are larger in \emph{env} and the wgs region, and smaller when restricting to only the tip branches. In some cases, the consensus phylogenies provide an adequate summary of the distribution (as in Sanger consensus phylogeny for all branches in \emph{env}, or the NGS consensus phylogeny for tip branches in \emph{env}). In other cases, the consensus phylogenies have estimates that are outliers in the distribution (as in both consensus phylogenies for either all or tip branches in the prrt region).

Taken together, these results demonstrate that within-host diversity is associated with variation in phylogenetic estimates, and that the consensus approach provides only a point estimate that does not always adeuqatley summarize the underlying variation. For example, the variation can be multi-modal and the point estimates from the consensus phylogenies can be outliers.

@@

Finally, we examine the clusters that are detected in the NGS vs. Sanger consensus phylogenies, and across the four regions. The NGS consensus phylogeny detects 8 clusters in prrt, 9 in \emph{int}, 4 in \emph{env}, and 6 in wgs. Each cluster appears as a colored bar in the phylograms in Figure 4. We use the proportion of times a cluster appears across the 500 profile-sampled phylogenies as a measure of support for that cluster, and report it next to the colored bar. Some clusters are robustly detected across all regions, such as the cluster (MC25, MC26, MC52), which has support between 98\% and 100\%. Other clusters are detected at lower support, and are not detected consistently across regions. For example, cluster (MC14, MC59) only appears in the wgs phylogeny, and at 62\% support.

The Sanger consensus phylogeny detects 8 clusters in prrt, 7 in \emph{int}, 6 in \emph{env}, and 7 in wgs (Figure 5). As with the NGS consensus phylogeny, the support and consistency of cluster detection varies across regions. Cluster detection is mostly consistent with the NGS consensus, except for the cluster (MC17, MC21), which only occurs in \emph{env} in the Sanger consensus, and has zero support from the profile samples. Three clusters are consistently detected across NGS and Sanger, and across all regions: (MC23, MC24); (MC25, MC26, MC52); and (MC27, MC28).

@@

Because previous studies of the HIV epidemic have shown high intra-patient variation, 
and because we estimate similar within-host diversity using HIV profiles of our NGS data,

@@

Our construction of HIV profiles from NGS data is limited by the accuracy of the NGS assays themselves. The codon frequencies in the profiles may be biased measures of the true within-host diversity because of biases in PCR amplification during sample preparation. Sequencing protocols such as Primer ID \parencite{jabara} have been introduced to reduce and correct for these biases. Unfortunately, the Primer ID protocol supports only a limited region of the HIV genome, and has not yet been extended to support near-whole genome sequencing of HIV, but future work could study this limited region to determine if cluster inference changes with the implementation of a Primer ID protocol.

The true HIV transmission network is unknown, but phylogenetic analysis and cluster inference are promising tools for aiding clinicians and public health officials \parencite{fauci}. Current phylogenetic approaches do not fully utilize the information on within-host diversity available in near-whole genome NGS data. As NGS data sets are increasingly available and more fully sample from the current HIV epidemic,
